Lost in Translation: \\ The function--sheaf dictionary for smooth commutative group schemes over finite fields

### 2014-04-10 dictionary.tex
Disappearing invisible quasicharacter sheaves
Car commit

\begin{abstract}
  We extend Deligne's function--sheaf dictionary from
  connected commutative algebraic groups over finite fields to smooth
  commutative group schemes over finite fields.
  To do so, we introduce the monoidal category of quasicharacter sheaves on
  smooth commutative group schemes $G$ over $\mathbb{F}_q$
  and show that the Trace of Frobenius is a functorial surjection
  from the group of isomorphism classes of quasicharacter sheaves on $G$
  onto the character group of $G(\mathbb{F}_q)$.
  If $G$ is connected, the kernel of this surjection is trivial.
  When $G$ is etale, however, the kernel is generally non-trivial,
  leading to interesting notion of invisible quasicharacter sheaves.
  We find a classification of invisible quasicharacter sheaves
  on etale commutative group schemes over $\mathbb{F}_q$
  in terms of the cohomology of $G(\mathbb{F}_q)$.
   \end{abstract}

@@

Although hidden in the discussion above, the connectness of $G$ played a crucial role in the geometrization and categorification of characters of $G(\Fq)$.  So it is natural to ask how the function--sheaf dictionary generalizes to not-necessarily connected commutative algebraic groups over finite fields. In this paper we answer that question by exhibiting a function--sheaf dictionary, in the sense above, for any smooth commutative group scheme $G$ over a finite field $\Fq$.
To do this we must pass from $\ell$-adic local systems on $G$ to Weil local systems on $G$
and also keep track of the isomorphisms \eqref{introbox} (mere existence of such an isomorphism is not enough), leading to the notion of {\it quasicharacter sheaves on $G$}.
%So, what is a quasicharacter sheaf?

A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms satisfying natural compatibilty conditions; here $\Frob{G}$ is the Frobenius automorphism of ${\bar G}$.

The main result of this paper is the following statement (see Theorem~\ref{thm:snake}):
the category $\QC(G)$ of quasicharacter sheaves on $G$
is a rigid monoidal category and Trace of Frobenius provides a functorial surjection
\begin{equation}
\QCiso{G} \to \Hom(G(\Fq), \EEx)
\end{equation}
with a canonical section,
and the kernel of the Trace of Frobenius is classified by $\Hh^1(\Weil{},\Hh^2(G(\bFq),\EEx))$.
In particular, in contrast to the case of connected commutative algebraic groups, the Trace of Frobenius may not injective.
We refer to quasicharacter sheaves in the kernel of Trace of Frobenius as \emph{invisible} quasicharacter sheaves.
The canonical section of the Trace of Frobenius allows one to convert any quasicharacter of $G(\Fq)$
into a quasicharacter sheaf on $G$.
In this way, quasicharacter sheaves geometrize characters of $G(\Fq)$.

@@

\subsection{Definitions}\label{sec:QCschemes}

@@

%\subsection{Main definition}\label{sec:category}

@@

We will see that the Trace of Frobenius may not provide complete
information about isomorphism classes of quasicharacter sheaves on $G$ when $G$ is not connected.
Our main tool for understanding this phenomenon is the function
\begin{equation}\label{2cocycle}
S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))
\end{equation}
defined here.
Let $\qcs{L} = (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$.
For each geometric point $x\in G(\bFq)$, choose a basis $\{ v_x \}$ for $\gqcs{L}_x$.
This choice determines a function $g : G(\bFq)\times G(\bFq) \to \EEx$ defined by
\[
\mu_{x,y}(v_{x+y}) = g(x,y) v_x \otimes v_y.
\]
Chasing through the diagram in Condition~\ref{QC.2} we see that
\[
g(x+y,z) g(x,y) = g(x,y+z) g(y,z),
\]
for all $x,y,z\in G(\bFq)$, so $g \in Z^2(G(\bFq),\EEx)$.
It is easy to see that the class of $g$ is independent of the choice made above;
let $\sigma_\mathcal{L}) \in H^2(G(\bFq),\EEx)$ the class of $g\in Z^2(G(\bFq),\EEx)$.
It is also easy to see that $S_G(\mathcal{L}) = S_G(\mathcal{L'})$ when $\qcs{L} \iso \qcs{L'}$. This defines
\[
S_G: \QCiso{G} \to H^2(G(\bFq),\EEx).
\]
We claim that $S_G(\qcs{L})$ is also fixed by the action of $\Weil{}$ on $H^2(G(\bFq),\EEx)$
inherited from the action of $\Weil{}$ on $G(\bFq)$.
To see this, we return to the choice $\{ v_x \in \gqcs{L}_x^\times \tq x\in G(\bFq) \}$
and define $f : G(\bFq) \to \EEx$ by
\[
\phi_x(v_{\Frob{G}(x)}) = f(x) v_x.
\]
Chasing through the diagram in Condition~\ref{QC.3} we see that
\begin{equation}\label{nohom}
g(\Frob{G}(x),\Frob{G}(y)) f(x) f(y)  = g(x,y) f(x+y)
\end{equation}
for all $x,y \in G(\bFq)$; thus, the class of $\,^{\Frob{}}g$ coincides with the class of $g$.
This concludes the definition of
$S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$.
It is easy to see that the function $S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$ is a group homomorphism:
\[
g_{\qcs{L}\otimes \qcs{L'}} = g_{\qcs{L}} \, g_\qcs{L'}
\]
for every $\qcs{L}, \qcs{L'} \in \QC(G)$.\todo{I still need to write down the details, for ourselves.}

@@

If $G$ is a connected and of finite type, then $\Hh^2(G(\bFq),\EEx)$ is trivial\todo{Reference?}
so all quasicharacter sheaves on connected commutative algebraic groups $G$ are visible;
 by contrast, if $G$ is etale then $\Hh^1(\Weil{},\Hh^2(G(\bFq),\EEx))$ may be non-trivial so $G$ may admit invisible quasicharacters sheaves.
We will return to that case in Section~\ref{sec:etale}.

@@

>>>>

\begin{abstract}
  We extend Deligne's function--sheaf dictionary from
  connected commutative algebraic groups over finite fields to smooth
  commutative group schemes over finite fields.
  To do so, we introduce the monoidal category of quasicharacter sheaves on
  smooth commutative group schemes $G$ over $\mathbb{F}_q$
  and show that the Trace of Frobenius is a functorial surjection
  from the group of isomorphism classes of quasicharacter sheaves on $G$
  onto the character group of $G(\mathbb{F}_q)$.
  If $G$ is connected, the kernel of this surjection is trivial.
  When $G$ is etale, however, the kernel is generally non-trivial,
  leading to the interesting notion of invisible quasicharacter sheaves.
  We find a classification of invisible quasicharacter sheaves
  on etale commutative group schemes over $\mathbb{F}_q$
  in terms of the cohomology of $G(\mathbb{F}_q)$.
  We also explain how to make invisible quasicharacter sheaves disappear,
  should the notion prove too troubling.
   \end{abstract}

@@

Although hidden in the discussion above, the connectness of $G$ played a crucial role in the geometrization and categorification of characters of $G(\Fq)$.  So it is natural to ask how the function--sheaf dictionary generalizes to not-necessarily connected commutative algebraic groups over finite fields. In this paper we answer that question by exhibiting a function--sheaf dictionary, in the sense above, for any smooth commutative group scheme $G$ over a finite field $\Fq$.
To do this we must pass from $\ell$-adic local systems on $G$ to Weil local systems on $G$
and also keep track of the isomorphisms \eqref{introbox}, leading to the notion of {\it quasicharacter sheaves on $G$}.

So, what is a quasicharacter sheaf?
A quasicharacter sheaf on $G$ is a triple $\qcs{L}\ceq
(\gqcs{L},\mu,\phi)$ where $\gqcs{L}$ is an $\ell$-adic local system on $\bG$ of rank~one and $\mu: m^*
\gqcs{L} \to \LxL$ and $\phi : \Frob{G}^*\gqcs{L} \to \gqcs{L}$ are isomorphisms satisfying natural compatibilty conditions; here $\Frob{G}$ is the Frobenius automorphism of ${\bar G}$.

The main result of this paper is the following statement (see Theorem~\ref{thm:snake}):
the category $\QC(G)$ of quasicharacter sheaves on $G$
is a rigid monoidal category and Trace of Frobenius provides a functorial surjection
\begin{equation}
\QCiso{G} \to \Hom(G(\Fq), \EEx)
\end{equation}
with a canonical section.
We refer to quasicharacter sheaves in the kernel of Trace of Frobenius as \emph{invisible} quasicharacter sheaves.
If $G$ is a connected commutative algebraic group then 
all quasicharacter sheaves on $G$ are visible.
In this paper we find a simple cohomological description 
of invisible quasicharacter sheaves on any smooth commutative
group scheme $G$ over $\Fq$.
In any case, the canonical section of the Trace of Frobenius allows 
one to convert any quasicharacter of $G(\Fq)$ 
into a quasicharacter sheaf on $G$;
consequently, quasicharacter sheaves geometrize and categorify 
characters of $G(\Fq)$.

@@

\subsection{Notations}\label{sec:QCschemes}

@@

\subsection{Main definition}\label{sec:category}

@@

We will see that the Trace of Frobenius may not provide complete
information about isomorphism classes of quasicharacter sheaves on $G$ when $G$ is not connected.
Our main tool for understanding this phenomenon is the function
\begin{equation}\label{2cocycle}
S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))
\end{equation}
defined here.
Let $\qcs{L} = (\gqcs{L},\mu,\phi)$ be a quasicharacter sheaf on $G$.
For each geometric point $x\in G(\bFq)$, choose a basis $\{ v_x \}$ for $\gqcs{L}_x$.
This choice determines a function $g : G(\bFq)\times G(\bFq) \to \EEx$ defined by
\[
\mu_{x,y}(v_{x+y}) = g(x,y) v_x \otimes v_y.
\]
Chasing through the diagram in Condition~\ref{QC.2} we see that
\[
g(x+y,z) g(x,y) = g(x,y+z) g(y,z),
\]
for all $x,y,z\in G(\bFq)$, so $g \in Z^2(G(\bFq),\EEx)$.
It is easy to see that the class of $g$ is independent of the choice made above;
let $\sigma_\mathcal{L}) \in H^2(G(\bFq),\EEx)$ the class of $g\in Z^2(G(\bFq),\EEx)$.
It is also easy to see that $S_G(\mathcal{L}) = S_G(\mathcal{L'})$ when $\qcs{L} \iso \qcs{L'}$. This defines
\[
S_G: \QCiso{G} \to H^2(G(\bFq),\EEx).
\]
We claim that $S_G(\qcs{L})$ is also fixed by the action of $\Weil{}$ on $H^2(G(\bFq),\EEx)$
inherited from the action of $\Weil{}$ on $G(\bFq)$.
To see this, we return to the choice $\{ v_x \in \gqcs{L}_x^\times \tq x\in G(\bFq) \}$
and define $f : G(\bFq) \to \EEx$ by
\[
\phi_x(v_{\Frob{G}(x)}) = f(x) v_x.
\]
Chasing through the diagram in Condition~\ref{QC.3} we see that
\begin{equation}\label{nohom}
g(\Frob{G}(x),\Frob{G}(y)) f(x) f(y)  = g(x,y) f(x+y)
\end{equation}
for all $x,y \in G(\bFq)$; thus, the class of $\,^{\Frob{}}g$ coincides with the class of $g$.
This concludes the definition of
$S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$.
It is easy to see that the function $S_{G}: \QCiso{G} \to H^0(\Weil{},H^2(G(\bFq),\EEx))$ is a group homomorphism.\todo{I still need to write down the details, for ourselves.}

@@

%If $G$ is a connected and of finite type, then $\Hh^2(G(\bFq),\EEx)$ is trivial\todo{Reference?} so all quasicharacter sheaves on connected commutative algebraic groups $G$ are visible; by contrast, if $G$ is etale then $\Hh^1(\Weil{},\Hh^2(G(\bFq),\EEx))$ may be non-trivial so $G$ may admit invisible quasicharacters sheaves. We will return to that case in Section~\ref{sec:etale}.

@@

\subsection{How to make invisible quasicharacter sheaves disappear}

Two approaches: subcategory and quotient category\todo{Under construction}
